---
title: Field Intelligence Patterns
author: "jl3205"
format: 
  html: 
    embed-resources: true
    toc: true
    code-fold: true
    theme: darkly
---

In this section, I examine the dataset through the lens of tactical intelligence — uncovering patterns across **time**, **geography**, **organizational presence**, **types of violence**, and **perpetrator characteristics**. Each visualization contributes to constructing a clearer picture of the **risk environment** faced by humanitarian actors in volatile regions.

*This EDA lays the groundwork for more advanced operational analysis — including anomaly detection, actor - motive clustering, and region - specific risk modeling.*

## Incident Hotspots by Geography
> How have incident locations shifted — and where do they concentrate most?
 
```{r, warning=FALSE}
suppressMessages(library(tidyverse))
suppressMessages(library(plotly))

df <- read_csv("clean_data/security_incidents_cleaned.csv", show_col_types = FALSE)

df_map <- df |> 
  filter(!is.na(Latitude), !is.na(Longitude)) |> 
  select(Year, Country, Latitude, Longitude, `Total affected`)

plot_geo(df_map) |>
  add_markers(
    x = ~Longitude, y = ~Latitude,
    frame = ~Year,
    size = ~`Total affected`,
    hoverinfo = "text",
    text = ~paste("Year:", Year, "<br>Country:", Country, "<br>Total affected:", `Total affected`),
    marker = list(color = 'red', opacity = 0.6, line = list(width = 0)),
    name = "Confirmed Incident Location"
  ) |>
  layout(
    margin = list(t = 90),
    title = "Evolving Threat Landscape: Aid Worker Attacks (1997–2025)",
    geo = list(
      projection = list(type = "natural earth"),
      showland = TRUE,
      landcolor = "gray10",
      countrycolor = "gray50",
      bgcolor = "#1e1e1e"
    ),
    paper_bgcolor = "#1e1e1e",
    plot_bgcolor = "#1e1e1e",
    font = list(color = "#FFFFFF"),
    legend = list(
    x = 0.9, 
    y = 0.9,
    font = list(size = 10)
  )
  )
```
> **Figure 2:** Animated progression of aid worker attacks from 1997 to 2025, revealing the volatility and shifting epicenters of operational risk across time.

```{r, warning=FALSE}
plot_geo(df_map) |>
  add_markers(
    x = ~Longitude, y = ~Latitude,
    size = ~`Total affected`,
    hoverinfo = "text",
    text = ~paste("Country:", Country, "<br>Total affected:", `Total affected`),
    marker = list(color = 'red', opacity = 0.5, line = list(width = 0))
  ) |>
  layout(
    margin = list(t = 90),
    title = "Cumulative Footprint of Aid Worker Attacks (1997–2025)",
    geo = list(
      projection = list(type = "natural earth"),
      showland = TRUE,
      landcolor = "gray10",
      countrycolor = "gray50",
      bgcolor = "#1e1e1e"
    ),
    paper_bgcolor = "#1e1e1e",
    plot_bgcolor = "#1e1e1e",
    font = list(color = "#FFFFFF"),
    showlegend = FALSE
  )
```
> **Figure 3:** High - density clusters in East Africa, South Asia, and the Sahel highlight persistent operational vulnerabilities — pinpointing regions where threat exposure is entrenched over time.

::: {.callout-note title="Insight: Urban-Rural Divide"}
Preliminary geospatial inspection suggests clustering in both urban flashpoints (e.g., Mogadishu, Kabul) and rural corridors (e.g., Sahelian belt), accentuating the challenge of one-size-fits-all security policies.
:::


## Organizational Risk Exposure
> Which agencies are most frequently involved in attacks?

```{r, warning=FALSE}
suppressMessages(library(plotly))

org_totals <- df |>
  summarise(
    UN = sum(UN),
    INGO = sum(INGO),
    ICRC = sum(ICRC),
    `NRCS and IFRC` = sum(`NRCS and IFRC`),
    NNGO = sum(NNGO),
    Other = sum(Other)
  ) |>
  pivot_longer(everything(), names_to = "Organization", values_to = "Incidents")

org_plot <- ggplot(org_totals, aes(x = reorder(Organization, -Incidents), y = Incidents)) +
  geom_col(
    aes(
      text = paste0(
        "Target: ", Organization, "<br>",
        "Confirmed Incidents: ", Incidents
      )
    ),
    fill = "#4EA8DE"
  ) +
  labs(
    title = "Organizational Exposure to Security Incidents",
    x = NULL,
    y = "Number of Incidents"
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))

ggplotly(org_plot, tooltip = "text") |>
  layout(
    margin = list(t = 90)
  )
```
> **Figure 4:** INGOs and NNGOs experience the brunt of violent incidents—suggesting that footprint scale, mission scope, and regional penetration correlate with heightened threat exposure.

## Victim Pathways: Aid Worker Roles and Incident Outcomes
> Who bears the brunt of violence - international or national staff?

```{r}
library(networkD3)
suppressMessages(library(htmlwidgets))

sankey_data <- df |> summarise(
  `National - Killed` = sum(`Nationals killed`, na.rm = TRUE),
  `National - Wounded` = sum(`Nationals wounded`, na.rm = TRUE),
  `National - Kidnapped` = sum(`Nationals kidnapped`, na.rm = TRUE),
  `International - Killed` = sum(`Internationals killed`, na.rm = TRUE),
  `International - Wounded` = sum(`Internationals wounded`, na.rm = TRUE),
  `International - Kidnapped` = sum(`Internationals kidnapped`, na.rm = TRUE)
) |> pivot_longer(everything(), names_to = "Path", values_to = "Count") |>
  separate(Path, into = c("Role", "Outcome"), sep = " - ")

nodes <- data.frame(name = unique(c(sankey_data$Role, sankey_data$Outcome)))

links <- sankey_data |> mutate(
  source = match(Role, nodes$name) - 1,
  target = match(Outcome, nodes$name) - 1,
  value = Count
) |> select(source, target, value)

sankey <- suppressMessages(sankeyNetwork(
  Links = links,
  Nodes = nodes,
  Source = "source",
  Target = "target",
  Value = "value",
  NodeID = "name",
  fontSize = 13,
  nodeWidth = 30,
  colourScale = JS("d3.scaleOrdinal().range(['#4EA8DE', '#FF5C5C', '#F4A261'])"),
  sinksRight = FALSE
))

htmlwidgets::onRender(sankey, "
  function(el) {
    function applyTheme() {
      const theme = document.documentElement.getAttribute('data-bs-theme');
      const color = (theme === 'dark') ? 'white' : 'black';
      d3.select(el).selectAll('.node text').style('fill', color);
    }

    // Apply on load
    applyTheme();

    // Re-apply when theme toggles
    const observer = new MutationObserver(applyTheme);
    observer.observe(document.documentElement, { attributes: true, attributeFilter: ['data-bs-theme'] });
  }
")
```
> **Figure 5:** National staff absorb the overwhelming majority of violence — particularly deaths and injuries — underscoring structural disparities in frontline exposure and security provisioning.


## Modus Operandi: Forms of Violence
What forms of violence are most common?
```{r, warning=FALSE}
attack_modes <- df |> count(`Means of attack`, sort = TRUE)

mod_plot <- ggplot(attack_modes, aes(x = n, y = reorder(`Means of attack`, n))) +
  geom_point(aes(text = paste0("Method: ", `Means of attack`, "<br>Incidents: ", n)),
             size = 4, color = "#FF5C5C") +
  labs(
    title = "Modus Operandi: Attack Types Against Aid Workers",
    x = "Reported Incidents", y = NULL
  ) +
  theme_minimal(base_size = 10)

ggplotly(mod_plot, tooltip = "text") |>
  layout(
    margin = list(t = 90)
  )
```
> **Figure 6:** Shootings, kidnappings, and bodily assaults dominate the landscape of violence — requiring field protocols that span both rapid - response and long - duration risk containment.

::: {.callout-warning title="Tactical Implication"}
Shootings and kidnappings account for the majority of attacks—suggesting field teams need flexible protocols that account for both short-range threats and prolonged abductions.
:::

## Threat Actor Profile
Who is behind these attacks?
```{r, warning=FALSE}
actor_types <- df |> count(`Actor type`, sort = TRUE)

actor_plot <- ggplot(actor_types, aes(x = n, y = reorder(`Actor type`, n))) +
  geom_segment(aes(x = 0, xend = n, yend = `Actor type`), color = "#F4A261") +
  geom_point(aes(text = paste0("Threat Actor: ", `Actor type`, "<br>Incidents: ", n)),
             size = 4, color = "#F4A261") +
  labs(
    title = "Threat Actor Landscape",
    x = "Confirmed Incidents", y = NULL
  ) +
  theme_minimal(base_size = 10)

ggplotly(actor_plot, tooltip = "text") |>
  layout(
    margin = list(t = 90)
  )
```
> **Figure 7:** Over **50%** of incidents list the perpetrator as “Unknown,” reflecting both fog-of-war conditions and the limitations of ground - level intelligence gathering in active zones.